### 2013-05-25 context-dependence.tex
tasep
more writeup.

The covariance calculations show that
if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
then $N(x,y)$ is close to Poisson.

Using this observation, we can estimate parameters by composite likelihood,
using all $N(x,y)$ as data.
Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.

To get confidence intervals, we need to either simulate, or do something different.

@@

@@

Modern techniques in sparse matrix computation (e.g.\ Krylov methods) provide efficient ways to do this
by only multiplying $G$ and $U$ some number of times, 
which is much more efficient because $G$ is sparse.

Also note that the further calculations in the phylogenetic case can be done using these methods.
To compute
\[
    \sum_z \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) 
    = \sum_z \left( e^{t_X G(m+2\ell)} \right)_{zx} \pi(z) \sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy} ,
\]
we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
multiply by $\pi(z)$, and then apply the algorithm again using the transpose of $G(m+2\ell)$.

Using the R package expm, this makes computation quite feasible:
with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively.

>>>>

The covariance calculations show that
if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
then $N(x,y)$ is close to Poisson.
Using this observation, we can estimate parameters by composite likelihood,
using all $N(x,y)$ as data.
Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.
However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.

@@

Here is one way to obtain such a set.
Take a pair of patterns $(x,y)$, that overlap at $m$ positions,
and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
For example, in the example above with $m=3$ and $\ell=2$,
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  C  & A & C & T & C & A & C \\
 $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$  
\end{tabular}
\end{center}
$x$ and $y$ differ only at (relative) position $i_1=2$.
Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
As another example,
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  A & C & T & C & A & C & G \\
 $y$ &  $\centerdot$  & $\centerdot$  & G & C & A & $\centerdot$  & $\centerdot$  
\end{tabular}
\end{center}
has $i_1 = \bar i = 1$.
No two such patterns can be shifted versions of each other
without adding more changed sites,
since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.

@@

Modern techniques in sparse matrix computation (Krylov methods) provide efficient ways to do this
by only multiplying $G$ and $U$ some number of times, 
which is much more efficient because $G$ is sparse.

\subsection{Updating $G$}

We can also take advantage of the sparsity of $G$ to perform efficient computation of the likelihood
under many sets of parameters.
First note that $G(m+2\ell)$ has, assuming only single-position changes,
$(1+m(|\calS|-1)) |\calS|^{m+2\ell}$ nonzero entries, since each of the $|\calS|^{m+2\ell}$ can change in $m$ places.
This will determine how the computation scales with $m$, $\ell$, and $|\calS|$,
once we precompute a number of things.
Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G$, in some fixed ordering;
sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
Each $g_i$ is a linear combination of mutation rates $\mu_j$, say $g_i = \sum_j a_{ij} \mu_j$;
so by precomputing $a_{ij}$ we can efficiently update $G$.
This is true if the boundary conditions are circular, mean-value, or neither.

\subsection{Phylogenetic peeling}

Furthermore, note that the further calculations in the phylogenetic case can be done using these methods.
To compute
\[
    \sum_z \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) 
    = \sum_z \left( e^{t_X G(m+2\ell)} \right)_{zx} \pi(z) \sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy} ,
\]
we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
multiply by $\pi(z)$, and then apply the algorithm again using the transpose of $G(m+2\ell)$.

Using the R package expm, this makes computation quite feasible:
with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
(and much longer for the whole matrix $e^{t G(8)}$).

%%%%%%
\section{Bayesian things}

So, suppose we've got a list of patterns $(x_i,y_i)$
and we're happy with the idea that they're independent.
Recall that
\[
\E\left[ N(x,y) \mid N(x) \right] = N(x) p_{m,\ell}(x,y) ,
\]
Parameters $\theta$ enter implicitly through $p_{m,\ell}$.

In the case that we observe the initial sequence, $N(x,y)$ are binomial with sample size $N(x)$.
and so the likelihood is 
\[
\prod_x N(x)! \prod_y \frac{ p_{m,\ell}(x,y)^{N(x,y)} }{ N(x,y)! } ,
\]
and so the negative log likelihood is, up to a constant,
\[
\sum_{x,y} N(x,y) \log p_{m,\ell}(x,y) .
\]
